Draft: Notes on the revised Hidden weight function

### 2024-10-25 RevisedHiddenWeight.tex
start modif part 2

@@

%\begin{definition}[Algebraic Immunity] \label{def:ai}
%	The algebraic immunity of a Boolean function $f\in \Bfn{n}$, denoted as $\AI(f)$, is defined as:
%	\[ \AI(f) = \min_{g \neq 0}\{ \degg(g) \; | \; fg = 0 \; \text{or} \; (f + 1)g = 0 \}{,} \]
%	where $\degg(g)$ is the algebraic degree of $g$.
%	The function $g$ is called an annihilator of $f$ (or $f + 1$). 
%	Additionally we denote $\AN(f) = \min_{g \neq 0}\{ \degg(g) \; | \; fg = 0\}$.
%\end{definition}
\begin{definition}[Nonlinearity and weightwise nonlinearity] \label{def:nl}
	The nonlinearity $\NL(f)$ of a Boolean function $f\in \BN$, where $n$ is a positive integer, is the minimum Hamming distance between $f$ and all the affine functions in $\BN$:
	\[ \NL(f) = \min_{g,\, \degg(g)\leq 1} \{ \hd(f,g) \}{,} \]
	where $g(x)=a\cdot x+\varepsilon$, $a\in \F_2^n, \varepsilon\in \F_2$. 	
	The nonlinearity can also be defined from the Walsh transform:
	\[ \NL(f) = 2^{n-1}- \frac{1}{2} \max_{a\in \F_2^n}{|\wt{f}(a)|}. \]
\end{definition}

@@

The $n$-variable \emph{Boolean symmetric functions} are those that are constant on each slice $\Ekn{k}{n}$ for $k\in [0,n]$. 
This class has been assiduously studied in the context of cryptography, see \eg \cite{IEEE:Carlet04,IEEE:CanVid05,INDO:BraPre05,DM:SarMai07,IEEE:QFLW09,Latin:Meaux19,CCDS:Meaux21,IEEE:CarMea21}.
In this paper we mainly consider the slice indicator functions.

%\begin{definition}[Elementary symmetric functions]
%Let $i\in [0,n]$, the elementary symmetric function of degree $i$ in $n$ variables, denoted $\sigma_{i,n}$, is the function which ANF contains all monomials of degree $i$ and no monomials of other degrees. 
%\end{definition}
\begin{definition}[Slice indicator functions]\label{def:slice}
 Let $k\in [0,n]$, the indicator function of the slice of weight $k$ is defined as:
 \[\forall  x\in \mathbb{F}_2^{n}, \quad \phikn{k}{n}(x) = 1 \text{ if and only if } \w(x) = k.\]
\end{definition}

%We recall that bent functions are those Boolean functions achieving maximal nonlinearity.

\pmnote{adapt usefull parts:

In~\cite{DAM:GinMea22} the authors study the relationship between weightwise perfectly balanced functions (functions balanced on all the slices, see for example~\cite{TOSC:CarMeaRot17,DCC:LiuMes19,CC:TangLiu19,BFA:MesSuLi21,ACI:ManKuLar22,SSRN:ZLCQZ23,ASIACCS:YCL+23,LATIN:GinMea23}) and weightwise affine functions \ie weightwise degree-$1$ functions. 
In fact weightwise degree-$1$ and degree-$0$ functions have been studied for their cryptographic properties in many works, without the same formalism. 
The weightwise constant functions ($\WD{0}{n}$) are the symmetric functions that have been extensively studied (\eg \cite{IEEE:Carlet04,IEEE:CanVid05,INDO:BraPre05,DM:SarMai07,IEEE:CheLu11,A2C:CarMea19,Latin:Meaux19,CCDS:Meaux21,IEEE:CarMea21}). 
Thereafter, the hidden weight bit function introduced in~\cite{IEEE:Bryant91} is a weightwise degree-$1$ function, the one obtained by fixing $f_0=0$ and $f_k=x_k$ for $k \in [n]$. 
The cryptographic properties of this function have been studied in~\cite{DAM:WCST14}, showing good algebraic properties for this function. 
In~\cite{TOSC:CarMeaRot17}, the bent functions in Propositions $1$ and $2$ are weightwise affine.}

\pmnote{Introduce formally HWBF, and weightwise degree d functions}
\begin{definition}[Hidden Weight Bit Function (HWBF)]\label{def:hwbf}
	Let $n\in \N^*$we call Hidden Weight Bit Function (HWBF) the $n$-variable Boolean function defined by:
	\[h(x)= \left\{ 
	\begin{array}{l l}
	0 & \quad \text{ if } \w(x) =0, \\
	x_k & \quad \text{ if } \w(x)=k.
	\end{array} \right. \]

@@

\begin{definition}[Weightwise degree-$d$ functions]\label{def:WdegD}
	Let $n\in \N^*$ and for $k\in[0,n]$ $\phikn{k}{n}$ denotes the indicator function of $\Ekn{k}{n}$. 
	An $n$-variable Boolean function $f$, written as $f=\sum_{k=0}^n f_k \phikn{k}{n}$, is called weightwise degree-$d$ if and only if for each $k\in [0,n]$ $f_k$ coincide with a function of degree at most $d$ over $\Ekn{k}{n}$. 
	%	Equivalently, $f$ is weightwise affine if and only if for each $k\in [0,n]$ $\NLk{k}(f)=0$.

	The set of weightwise degree-$d$ functions is denoted by $\WD{d}{n}$.

	%	$f=\sum_{k=0}^n f_k \phikn{k}{n}$ is called weightwise affine if and only if for each $k\in [0,n]$ $f_k$ coincide with an affine function over $\Ekn{k}{n}$.

\pmnote{add: properties of maj function, HWPF, and some WWL and WWQ}

\pmnote{add the nonlinearity of Dickson's function}

@@

\section{Main section}
%\subsection{Restricted Walsh transform of $f$ in a recursive fashion}

We study the function $f$ in an even number of variables $n$ defined as the following:
\[f(x_1,\cdots,x_n)=\left(\sum_{i=1}^{n/2} (x_i+1) x_{i+\frac{n}{2}}\right) + \sum_{k=1}^n \phikn{k}{n} x_k.\]

>>>>

\pmnote{short intro:

HWBF introduced in x, interesting for its cryptographic criteria and easy implementation. cryptographic properties studied in y such as balancedness, algebraic immunity and nonlinearity. 
New potential uses in a stream cipher, or modifications. Use: transciphering and works aiming at finding Boolean functions easy to evaluate homomorphically, input oblivious algorithms. 
Nonlinearity a bit low for direct use.

Contributions: new function as addition of HWBF and quadratic bent function. We show when the function is balanced, formalize and study its nonlinearity. Bounds on the nonlinearity of $f$, generalization to a family of function, and an algorithm for estimating the nonlinearity for $n$ up to $80$. Comparisons with other functions and practical values of the parameters for $n$ up to $20$.

}

@@

%\begin{definition}[Algebraic Immunity] \label{def:ai}
%	The algebraic immunity of a Boolean function $f\in \Bfn{n}$, denoted as $\AI(f)$, is defined as:
%	\[ \AI(f) = \min_{g \neq 0}\{ \degg(g) \; | \; fg = 0 \; \text{or} \; (f + 1)g = 0 \}{,} \]
%	where $\degg(g)$ is the algebraic degree of $g$.
%	The function $g$ is called an annihilator of $f$ (or $f + 1$). 
%	Additionally we denote $\AN(f) = \min_{g \neq 0}\{ \degg(g) \; | \; fg = 0\}$.
%\end{definition}
\begin{definition}[Nonlinearity %and weightwise nonlinearity
	] \label{def:nl}
	The nonlinearity $\NL(f)$ of a Boolean function $f\in \BN$, where $n$ is a positive integer, is the minimum Hamming distance between $f$ and all the affine functions in $\BN$:
	\[ \NL(f) = \min_{g,\, \degg(g)\leq 1} \{ \hd(f,g) \}{,} \]
	where $g(x)=a\cdot x+\varepsilon$, $a\in \F_2^n, \varepsilon\in \F_2$. 	
	The nonlinearity can also be defined from the Walsh transform:
	\[ \NL(f) = 2^{n-1}- \frac{1}{2} \max_{a\in \F_2^n}{|\wt{f}(a)|}. \]
\end{definition}

When $n$ is even, the nonlinearity of a function can reach at most $2^{n-1}-2^{n/2 -1}$. Functions reaching this maximum are called bent functions. The quadratic function $d_n$ given by: $d_n(x)=\sum_{i \text{ odd}}^n x_i x_{i+1}$ is an example of bent function.

@@

The $n$-variable \emph{Boolean symmetric functions} are those that are constant on each slice $\Ekn{k}{n}$ for $k\in [0,n]$. 
This class has been assiduously studied in the context of cryptography, see \eg \cite{IEEE:Carlet04,IEEE:CanVid05,INDO:BraPre05,DM:SarMai07,IEEE:QFLW09,IEEE:CheLu11,Latin:Meaux19,CCDS:Meaux21,IEEE:CarMea21}.
In this article we mainly consider the slice indicator functions.
%\begin{definition}[Elementary symmetric functions]
%Let $i\in [0,n]$, the elementary symmetric function of degree $i$ in $n$ variables, denoted $\sigma_{i,n}$, is the function which ANF contains all monomials of degree $i$ and no monomials of other degrees. 
%\end{definition}
\begin{definition}[Slice indicator functions]\label{def:slice}
 Let $k\in [0,n]$, the indicator function of the slice of weight $k$ is defined as:
 \[\forall  x\in \mathbb{F}_2^{n}, \quad \phikn{k}{n}(x) = 1 \text{ if and only if } \w(x) = k.\]
\end{definition}

Bigger families of function can be obtained by considering functions of bounded degree on each slice, it corresponds to the concept of weightwise degree-$d$ function introduced in~\cite{DAM:GinMea22} for $d=1$ and~\cite{DAM:MeaOza24} for the general case.

\begin{definition}[Weightwise degree-$d$ functions (~\cite{DAM:MeaOza24} Definition 12)]\label{def:WdegD}
	Let $n\in \N^*$ and for $k\in[0,n]$ $\phikn{k}{n}$ denotes the indicator function of $\Ekn{k}{n}$. 
	An $n$-variable Boolean function $f$, written as $f=\sum_{k=0}^n f_k \phikn{k}{n}$, is called weightwise degree-$d$ if and only if for each $k\in [0,n]$ $f_k$ coincide with a function of degree at most $d$ over $\Ekn{k}{n}$. 
		The set of weightwise degree-$d$ functions is denoted by $\WD{d}{n}$.
\end{definition}

Various weightwise affine functions (\ie belonging to $\WD{1}{n}$) have been exhibited such as in~\cite{TOSC:CarMeaRot17} where the bent functions in Propositions $1$ and $2$ are weightwise affine or in~\cite{DAM:GinMea22} to show than no weightwise perfectly balanced function is weightwise affine for $n\ge 8$. 
The most known example of weightwise affine function is the hidden weight bit function introduced in~\cite{IEEE:Bryant91}, the one obtained by fixing $f_0=0$ and $f_k=x_k$ for $k \in [n]$. The cryptographic properties of this function have been studied in~\cite{DAM:WCST14}, showing good algebraic properties for this function.

\begin{definition}[Hidden Weight Bit Function (HWBF)]\label{def:hwbf}
	Let $n\in \N^*$we call Hidden Weight Bit Function (HWBF) the $n$-variable Boolean function defined by:
	\[h(x)= \left\{ 
	\begin{array}{l l}
	0 & \quad \text{ if } \w(x) =0, \\
	x_k & \quad \text{ if } \w(x)=k.
	\end{array} \right. \]

@@

In~\cite{DAM:MeaOza24}, the parameters of different functions from $\WD{1}{n}$ and $\WD{2}{n}$ are studied experimentally for $n$ up to $20$ and lower bounds are given for the nonlinearity. 
We summarize the known results on the cryptographic criteria of weightwise affine functions we will use in this article in the following property.

\pmnote{complete the following prop, + references}

\begin{property}[Cryptographic properties of functions from $\WD{1}{n}$ \label{prop:wwd1}]
Let $n\in \N^*$, the following hold for functions in $\WD{1}{n}$.
\begin{itemize}
	\item Nonlinearity of majority functions.
	\item Nonlinearity of HWBF.
	\item Nonlinearity bound WWA.
\end{itemize}

\end{property}

@@

\section{Revisited HWBF and balancedness}
%\subsection{Restricted Walsh transform of $f$ in a recursive fashion}

In this part we formally define the revisited HWBF function $f$ and introduce a quantity that allows us to study its weightwise nonlinearity and balancedness.

\begin{definition}[Revisited HWBF]\label{def:revHWBF}
Let $n\in \N^*$, even, we call revisited HWBF and denote by $f$ the Boolean function defined as:
\[f(x_1,\cdots,x_n)=\left(\sum_{i=1}^{n/2} (x_i+1) x_{i+\frac{n}{2}}\right) + \sum_{k=1}^n \phikn{k}{n} x_k.\]
\end{definition}

%We study the function $f$ in an even number of variables $n$ defined as the following:
%\[f(x_1,\cdots,x_n)=\left(\sum_{i=1}^{n/2} (x_i+1) x_{i+\frac{n}{2}}\right) + \sum_{k=1}^n \phikn{k}{n} x_k.\]
